Increasing prevalence of plant-fungal symbiosis across two centuries of environmental change

### 2025-02-05 Manuscript/Reviewer_response_endo_herbarium.tex

\comment{While the research questions were clearly stated and clearly tested and answered by their results (clear and well structured), the hypothesis as well as the importance of the findings remains, in my opinion, “unbaked” at some points. Although the authors legitimately emphasized the importance and robustness of their data and methods throughout the manuscript, by contextualizing their results using recent literature and proposed some future research directions, the broader impact of this study on the field remains insufficiently addressed. A more in-depth discussion about the potential causes (ecological) of the results, and their implications for the ecology of the host plants or plants-endophytes interactions would be highly beneficial.}
\response{We appreciate Reviewer 2's positive evaluation of our work in regards to its success in answering our central research questions. Their review has helped us to thoroughly revise the structure of our work and improve organization. In addition, their desire to see more in-depth discussion of the ecological implications of our findings was echoed by Reviewer 1, and we have added new discussion material addressing this topic, detailed in response to specific comments.}
\linesref{ms-R2C14-begin}{ms-R2C14-end}

\comment{In the abstract, I recommend briefly mentioning the statistical analyses or framework used (e.g., INLA) before presenting the results to enhance readability.}
\response{\jacob{We have revised the abstract to include information about the statistical analysis.}{Please see my comment in the abstract}}
\linesref{ms-R2-C15-begin}{ms-R2-C15-end}

\comment{Introduction: While, there is nothing scientifically wrong, I believe that the Introduction can be significantly shortened by going straight to the point. I suggest removing un-necessary examples (i.e, corals or insects’ symbioses), and reorganizing the content for clarity. I found some back-and-forth between discussions on methodological issues (i.,e L49-L60 ; L112-L118)) and the ecological importance of endophytes for host plants (L40-43; L75-82; and vice-versa). I would recommend to group the paragraphs presenting the research gaps of the ecological importance of endophytes, particularly in the context of climate change together and addressing the methodological novelty of this study in a separate, cohesive paragraph.}
\response{\jacob{}{You forgot to answer this comment. I think we should push back. Here is my proposition: "We appreciate the suggestion to shorten the  Introduction and improve its clarity. While we understand the value of streamlining the content, we believe that the examples provided, including those related to coral and insect symbioses, serve to contextualize the study within broader ecological frameworks and may help engage readers who are less familiar with the topic."}\tom{}{I like Jacob's suggestion here and agree that we should push back on cutting the non-plant examples. But maybe the reorganization suggested here could work. I have not looked at the ms yet, maybe you have already done this.}}

\comment{Discussion: The Discussion is the most difficult section to follow. Similarly, to the Introduction, I back-and-forth between paragraphs/arguments. I believe that the use of sub-headings (in your rewriting before to remove them) would greatly improve clarity by following the same order as your research questions. For example, you could begin by discussing about the spatial/temporal patterns in prevalence, followed by the relationships with climatic drivers before to conclude with the technical/ecological novelty of your findings.}
\response{Thank you for the constructive suggestions. In response to this comment, we have added sub-headings to our Discussion and revised our discussion section for clarity and flow between ideas. Many of these changes are detailed in response to further comments from Reviewer 2.}

\comment{Supplementary figures: Fig.A3; A4; A5; A6;A7;A8;A9;A10; A11;Table A1. The captions for these figures do not provide enough information to understand their meaning, making the supplementary figures unexploitable as such. To assist readers (who would otherwise need to juggle between the supplementary figures and the main text), please ensure that each figure’s caption includes: the specific model or analysis from which it was produced; the data used (e.g., whether it is based on all host plants or a particular plant, which period of data used, etc.); and defining terms or acronyms of these figures (examples: A and T from figure A6; $\tau$ from A7; conservative vs liberal A8; AGHY vs ELVI A11…..).}
\response{We appreciate Reviewer 2's attention to our supplemental material. We have revised all of the figure captions in the supplement to provide the information that readers will need to interpret this material. These edits are also detailed in our responses to specific comments below.}

\comment{L54: Which physical legacies or ecological processes? Provide example.}
\response{In response to this comment, we have added additional information about the examples of ecological processes studied previously using herbarium specimens, \tom{such as...}{Again, be sure to answer questions here. It's ok (and expected) to have redundancy between the manuscript and response letter.}.}
\linesref{ms-R2C19-begin}{ms-R2C19-end}

\comment{L64-L74: Please remove example of symbiosis from other systems and focus on your targeted endophytic symbiosis. (Remove corals, insects or lichens.)}
\response{We appreciate the suggestion but, given the broad readership of \textit{Global Change Biology}, we strongly feel that these other symbioses provide relevant examples of the potential responses of symbioses to global change, and that they help place our work into a broader context. Our central point is that climate change responses are known in a few key symbiota, but not in the grass-endophyte symbiosis that is our focus, and that studies across diverse taxa are necessary to understand \emph{why} responses can tip from resilience to disruption of the interaction. These examples motivated our study and so we have opted to keep them in the text.}

\comment{-L104: In which sense? How this hypothesis would be reflected in your results? Would you expect a distribution shifts? Please specify.}
\response{We have edited this sentence to include a specific prediction for how this hypothesis would relate to our study system. We have also added specific predictions \tom{two paragraphs further down}{I would provide line numbers for this change.}, where we present our study's questions.}
\linesref{ms-R2C21-begin}{ms-R2C21-end}

\comment{-L131-134: Super nice data!}
\response{We thank this Reviewer 2 for acknowledging the value of our dataset, which represents significant time and effort.}

@@

\comment{-L140: Maybe trivia, so please feel free to ignore if I am wrong, but maybe specify that C3 refers to a metabolic type of plant?}
\response{Thank you for this suggestion. We have added additional context about C3 photosynthesis.}
\linesref{ms-R2C23-begin}{ms-R2C23-end}

\comment{-L167-168: What is the spatial resolution of the different units? What could be the distance between counties or between state centroids. And how big or small county can be?}
\response{In response to this comment, we have added information about the relative size of the bounding boxes generated during the geocoding process and the pairwise distance between points. Because we included coordinates georeferenced to local municipalities, counties, and in a few instances, states, the width of this bounding box varied from less than 1 km. to over 1233 km (median 841 km.).\tom{}{Good example of answering the question AND flagging the revised line number.}}
\linesref{ms-R2C24-begin}{ms-R2C24-end}

\comment{-L176: “Partial” would be more adapted than “imperfect”.}
\response{"Imperfect" transmission is a commonly use term within the study of symbiosis, therefore we have opted to retain its use for consistency with previous literature. We have, however, also added a definition of the term "imperfect transmission" at the point it is first used.}
\linesref{ms-R2C25-begin}{ms-R2C25-end}

\comment{-L180-187: I found that the use of conservative and liberal very confusing. And I am not sure I understood properly what they are. Moreover, I you do not mention these type of status in the results nor discussion. Please reformulate and clarify in the main text and maybe it can be used in the discussion?}
\response{We are sorry that this terminology has caused confusion, and we appreciate the reviewer bringing this to our attention. In response to this feedback, we have made several edits to provide additional explanation. 
We record two scores for each sample to evaluate uncertainty in the process of assigning endophyte status under the microscope. In cases with unambiguous identification (presence or absence), the two scores are the same. When there is uncertainty in identification, we assign a positive ``liberal'' score and a negative ``conservative'' score. \tom{}{I would add here that ``split'' scores are infrequent, and give the frequency.} We present a comparison of results based on this scoring choice in our supplement, which showed minimal difference, and therefore we only present the results of the "liberal" score for the rest of the central results.}
\linesref{ms-R2C26-begin}{ms-R2C26-end}

@@

\comment{-L190: Spatial resolution of the aggregation?}
\response{Our global slope estimates a trend averaging across the entire study region. We have changed this to "averaged" instead of "aggregated". We have added additional information elsewhere about the spatial resolution of our analysis in response to Reviewer 2's subsequent comments.}
\linesref{ms-R2C27-begin}{ms-R2C27-end}

\comment{-L215: “borrow information” is more appropriate than “borrow strength”.}
\response{Thank you for pointing this out. We have made this change.}
\linesref{ms-R2C28-begin}{ms-R2C28-end}

\comment{-L229: Why 5? Is there a reference?}
\response{We have added a reference supporting the use of penalized complexity priors, and revised terminology to more closely reflect the parameter values used to implement the model in inlabru.}
\linesref{ms-R2C29-begin}{ms-R2C29-end}

\comment{-L241-L258: Why this paragraph appears here? I understand that this is for the spatial distribution of the endophytes (within a host) but should it be in the first paragraph? Should it be treated right after the prevalence, and before to mention the spatial models? I feel like this paragraph is coming out of the blue and is making the full material and methods confusing.}
\response{We appreciate the push to improve the organization of our manuscript. In this case, we see the current placement of the section "Modeling distributions of host species" as appropriate because the preceding section describes models of endophyte prevalence that are independent from these models of host species' distributions. The predictions from the host distribution models are used principally for identifying realistic geographic ranges over which we test the relationship between climate drivers and trends in endophyte prevalence, and for visualizing relationships across study region. In response to this comment, we have added additional context describing this role that the host distribution models play.\tom{}{Maybe you say this in the text but I would point out that the herbarium records do not reflect the entire species' ranges, so if we want to make range-wide predictions we need some other way to do this.}}
\linesref{ms-R2C30-begin}{ms-R2C30-end}

\comment{-L281: “which” instead of “whether”?}
\response{This change would alter the meaning of the question posed in this sentence, which we feel would not accurately reflect our intention. We have opted to keep "whether".}

@@

\comment{-L285-288: Spatial resolution?}
\response{We now provide the spatial resolution of the pixels generated. We have also added added explanation that the choice of pixel number/size is arbitrary, relevant to Reviewer 2 Comment 37.}
\linesref{ms-R2C32-begin}{ms-R2C32-end}

\comment{-L316: “Confidence” instead of “certainty”}
\response{We have made this change. Thank you.}
\linesref{ms-R2C33-begin}{ms-R2C33-end}

\comment{-Figure 2: Very nice figure.}
\response{We thank reviewer 2 for this supportive comment.}

\comment{-L323: “We found no evidence that collector biaises influenced our results” contradictory to L325 “The identity of individual scorers did contribute to observed patterns in endophyte prevalence.” I am confused. They are quite contradictory no?}
\response{Reviewer 2 confuses 'collector', those researchers who collected herbarium specimens, with the individuals we term 'scorer', those members of our lab who performed the microscope work to identify \emph{Epichloë} fungal endophytes. 
In response to this comment, we have more clearly defined 'collector' and 'scorer' in our manuscript.
Aside from this confusion, there is no contradiction in our results; we found that collector identify did not influence model predictions, while scorer identity did influence model predictions, as evidenced in Figs. A9-A10. }
\linesref{ms-R2C35-begin}{ms-R2C35-end}
\linesref{ms-R2C35-b-begin}{ms-R2C35-b-end}

@@

\comment{L328-333: This section should be in discussion rather than in results no?.}
\response{We have moved this interpration of the impact of scorer and collector random effects into our discussion section.}
\linesref{ms-R2C36-begin}{ms-R2C36-end}

\comment{-Figure 4: The model that have been used to produce these relationships should be mentioned in the caption. Moreover, in the caption, “points of 250 randomly sampled pixels across each host distribution”. First: Why not taking more points/all points? And what are the spatial resolution of these pixels?}
\response{ Our model uses a continuous spatial process to model variation across space in the estimated trends in endophyte prevalence. This permits us to generate predictions from the model at any location across the study region at high resolution, however doing so does require longer computation time. Models fit to the full set of pixels for each species show climate regressions in a similar direction as shown here and have narrower credible intervals. We could, for example, generate many thousands of points across each distribution, but \tom{this would make the credible intervals largely meaningless}{By this logic the CIs are meaningless for any sample size, whether its the full range or an arbitrary subset.}. We have added material describing why the choice of pixel dimensions is arbitrary as described in response to Reviewer 2 Comment 32. We also note in our results section that these relationship do not differ when using all points (Line \ref{ms-R2C37-begin}). In addition, we have edited the figure caption to include information about the origin of model (Now Figure 5).}
\linesref{ms-R2C37-begin}{ms-R2C37-end}
\linesref{ms-R2C32-begin}{ms-R2C32-end}

@@

\comment{- L382-386: This should be in discussion rather than in results.}
\response{In response to this comment, we have edited the sentence referenced in this comment to remove interpretation and focus solely on results. We feel the sentence referenced in this comment is appropriately placed in that it presents the magnitude of the climate relationship for \emph{E. virginicus}, similarly to each of the other two host species.}
\linesref{ms-R2C38-begin}{ms-R2C38-end}

\comment{L387-398: I feel like this part is coming quite late in the results. You have already been presented the results of the models, and relationships, I would suggests to move this earlier in the results section. Starting from model performance before to show results produced from this model.}
\response{We appreciate the suggestion but we respectfully disagree that this change would improve the organisation of our manuscript. As currently organized, the introduction presents questions that explore 1) trends in Epichloë prevalence, 2) the climate drivers of these trends, and 3) evaluating models to improve global change forecasts. Following this and other comments from Reviewer 2, we have edited to ensure that this organization of topics is repeated in the presentation of methods, in results, and in discussion, and added subheadings to better organize our manuscript.}

@@

\comment{L453-455: This part is hard to follow, please rephrase?}
\response{We have edited these sentences for clarity.}
\linesref{ms-R2C43-begin}{ms-R2C43-end}

\comment{L474: This is a good point that should maybe put in relation with other global changes such as land use changes. This would be nice to emphasize that probably land uses change impacted a lot/or not the natural populations of the three host species and emphasize that could also have impacted the ecological dynamics of your symbiosis at larger scale?}
\response{We thank Reviewer 2 for this interesting observation. We have in fact been working on a separate paper that analyzes the influence of land use changes and other anthropogenic disturbances on endophyte symbiosis. In response to this comment, we have added additional material to our discussion that proposes the potential potential for these other global change drivers to impact the dynamics of the symbiosis.}
\linesref{ms-R2C44-begin}{ms-R2C44-end}

\comment{L498: This is a good point as well. Moreover, the uses of eDNA metabarcoding or other sequencing methods to characterise the identity of your Epichloe composition would be very interesting to disscuss at this point. Would you expect a turnover of species (from Epichloe genera), that could impact the performance and potentially something about the dynamics of the plant-symbiotic dynamic?}
\response{This is an interesting question. \tom{}{I would add a brief response to the question here, even a quick summary of whatever speculation you added to the Discussion.}In response to this comment, we have expanded our discussion of the potential influence that host-symbiont genotype may have on our focal species' range dynamics in our re-organized discussion.}
\linesref{ms-R2C45-begin}{ms-R2C45-end}

@@

\comment{L501: “Second”, I did not see the “First”.}
\response{We have fixed this.}
\linesref{ms-R2C46-begin}{ms-R2C46-end}

\comment{Supplementary figures:-Page 1; I don’t know if its journal guidelines, but, should there be a title page, with authors information, affiliations and a title being Appendix to “Increasing prevalence of plant-fungal symbiosis across two centuries of environmental change” ?}
\response{We do not believe this is a journal requirement, however in response to this comment, we have added the requested information at the start of our supplementary material to connect clearly its content to our manuscript.}
\linesref{ms-R2C47-begin}{ms-R2C47-end}

\comment{Fig.A3: Which data were used to produce the simulated datasets? Which model is this from?}
\response{The "simulated data" here is generated as a posterior predictive test of our model. We predicted new values of endophyte occurence for each row in our observed dataset from model fitted to herbarium data. We have edited this for clarity, instead referring to "predicted" vs "observed" data in the figure caption (Figure A3), and have added a relevant citation within our methods.}
\linesref{ms-R2C48-begin}{ms-R2C48-end}

\comment{A4 and A5: Which model, which data? (All plants, all years)?}
\response{We have added description of the model and data for these supplemental figures (A4 and A5).\tom{}{And respond to the question here.}}

@@

\comment{A6; From which models this is from? How probabilities are calculated? What is A and T?}
\response{We have added description of the model that generates these parameter estimates in Fig. A6. We have also provided additional explanation of the parameters A and T, which come from Eqn. 1 describing the endophyte prevalence model.}

\comment{A7; From which models; what is the legend for the value means, what is $\tau$?}
\response{We have added description of the model that generates these parameter estimates in Fig. A7. We have also provided additional explanation of the parameter  $\tau$, which come from Eqn. 1 describing the endophyte prevalence model.}

\comment{A8;redefine what is liberal vs conservative; what AGPE, AGHY and ELVI means if the Panel B-C-F is from which model. What are the value? (This is an average per year, across 200 years?)}
\response{In response to this comment, we have redefined liberal vs. conservative scores in the figure caption of Fig. A8, along with adding information that describes the model from which these parameter estimates are derived. We have re-plotted the figure with species names in place of the four letter codes noted by Reviewer 2.}
